---
title: "Reproducible Research Assignment 4"
author: "SitiSoraya"
date: "January 14, 2020"
output: html_document
---

## Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
```

## Data processing
Reading data using read.csv method

```{r}
storm_data <- read.csv("repdata_data_StormData.csv")
head(storm_data)
```

Extract relevant columns for processing
```{r}
storm <-storm_data[,c(8,23:28)]
rm(storm_data)
head(storm)
```

## Summary of fatalities and injuries
```{r}
summary(storm$FATALITIES)
summary(storm$INJURIES)
```

## Total injuries by event type (top 20 by descending order) -- using aggregate function
```{r}
tot_injuries <- aggregate(INJURIES~EVTYPE, storm, sum)
tot_injuries <- arrange(tot_injuries, desc(INJURIES))
tot_injuries <- tot_injuries[1:20,]
tot_injuries
```

## Total fatalities by event type (top 20 by descending order) -- using aggregate function
```{r}
tot_fatalities <- aggregate(FATALITIES~EVTYPE, storm, sum)
tot_fatalities <- arrange(tot_fatalities, desc(FATALITIES))
tot_fatalities <- tot_fatalities[1:20,]
tot_fatalities
```

## Results
```{r}
totals <- merge(tot_fatalities, tot_injuries, by.x = "EVTYPE", by.y = "EVTYPE")
totals <- arrange(totals, desc(FATALITIES+INJURIES))
event_names <- totals$EVTYPE
barplot(t(totals[,-1]), names.arg = event_names, ylim = c(0,95000), beside = T, cex.names = 0.8, las = 2, col = c("yellow", "purple"), main = "Top Disaster Casualities")
legend("center", c("Fatalities","Injuries"), fill = c("yellow", "purple"), bty = "n")
```

## Data processing
Convert PROPDMGEXP into numbers, for e.g., H = 10^2 and create new variable called PROPDAMAGE
```{R}
storm$PROPDAMAGE = 0
storm[storm$PROPDMGEXP == 'H',]$PROPDAMAGE = storm[storm$PROPDMGEXP == "H",]$PROPDMG * 10^2
storm[storm$PROPDMGEXP == 'K',]$PROPDAMAGE = storm[storm$PROPDMGEXP == "K",]$PROPDMG * 10^3
storm[storm$PROPDMGEXP == 'M',]$PROPDAMAGE = storm[storm$PROPDMGEXP == "M",]$PROPDMG * 10^6
storm[storm$PROPDMGEXP == 'B',]$PROPDAMAGE = storm[storm$PROPDMGEXP == "B",]$PROPDMG * 10^9
```

Convert CROPDMGEXP into numbers, and create new variable called CROPDAMAGE
```{R}
storm$CROPDAMAGE = 0
storm[storm$CROPDMGEXP == 'H',]$CROPDAMAGE = storm[storm$CROPDMGEXP == "H",]$CROPDMG * 10^2
storm[storm$CROPDMGEXP == 'K',]$CROPDAMAGE = storm[storm$CROPDMGEXP == "K",]$CROPDMG * 10^3
storm[storm$CROPDMGEXP == 'M',]$CROPDAMAGE = storm[storm$CROPDMGEXP == "M",]$CROPDMG * 10^6
storm[storm$CROPDMGEXP == 'B',]$CROPDAMAGE = storm[storm$CROPDMGEXP == "B",]$CROPDMG * 10^9
```

Aggregate property and crop damage into a variable called damage, then arrange by descending order (top 20)
```{r}
damage <- aggregate(PROPDAMAGE + CROPDAMAGE ~ EVTYPE, storm, sum)
names(damage) = c("EVENT_TYPE", "TOTAL_DAMAGE")
damage$TOTAL_DAMAGE <- damage$TOTAL_DAMAGE/10^9
damage$EVENT_TYPE <- factor(damage$EVENT_TYPE, levels = damage$EVENT_TYPE)
damage <- arrange(damage, desc(TOTAL_DAMAGE))
damage <- damage[1:20,]
head(damage)
```

```{r}
with(damage, barplot(TOTAL_DAMAGE,names.arg = EVENT_TYPE, beside = T, cex.names = 0.8, las = 2, col = "blue", main = "Top 20 Event Type of Property and Crop Damage", ylab = "Total Damage in USD (10^9"))
```